---
title: "What are the best times and days of the week to minimise delays each year?"
author: 'Vishma Arachchi'
date: "2024-02-05"
output: html_document
---

What are the best times and days of the week to minimise delays each year?



LOADING LIBRARIES

```{r}

library(ggplot2)
library(repr)
library(tidyr)
library(dplyr)
library(reshape2)


```

LOADING DATASET

```{r}
df_a <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/2006.csv.bz2')

```

Checking for missing values and empty strings

```{r}
df_a[df_a == ""] <- NA

missing_values <- colSums(is.na(df_a))
print(missing_values)

```

Conversion of columns to type numeric


```{r}
df_a$DepDelay <- as.numeric(df_a$DepDelay)
df_a$ArrDelay <- as.numeric(df_a$ArrDelay)
df_a$CRSDepTime <- as.character(df_a$CRSDepTime)
```



Calculating the mean of ArrDelay and DepDelay


```{r}
mean_ArrDelay <- mean(df_a$ArrDelay,na.rm = TRUE)
mean_DepDelay <- mean(df_a$DepDelay,na.rm = TRUE)
```

Replacing NA values


```{r}

df_a$ArrDelay[is.na(df_a$ArrDelay)] <- round(mean_ArrDelay,1)
df_a$DepDelay[is.na(df_a$DepDelay)] <- round(mean_DepDelay,1)

```

Finding the correlation between arrival delays and departure delays()

```{r}
correlation_coefficient <- cor(df_a$ArrDelay, df_a$DepDelay)
print(correlation_coefficient)
```

This is a very high value suggesting that a combined column is appropriate


New dataframe for relevant columns

```{r}
df_main <- select(df_a,DayOfWeek, CRSDepTime, ArrDelay, DepDelay )

```

Creating new column for Total Delays


```{r}
df_main$Total_Delays <- df_main$ArrDelay + df_main$DepDelay

```

Creating Dataframe where only datapoints with delays are included


```{r}
df_only_delays <- subset(df_main, Total_Delays > 0)

```

Adding leading zeros to the CRSDepTime Column


```{r}
df_only_delays$CRSDepTime <- as.character(df_only_delays$CRSDepTime)
df_only_delays$CRSDepTime <- sprintf("%04s", df_only_delays$CRSDepTime)
```



Extracting Hour of the day


```{r}
hours <- substr(df_only_delays$CRSDepTime, nchar(df_only_delays$CRSDepTime) - 3, nchar(df_only_delays$CRSDepTime) - 2)
hours <- as.numeric(hours)
df_only_delays$Hours <- hours
df_only_delays$CRSDepTime <- as.numeric(df_only_delays$CRSDepTime)

```


Find unique values in the column

```{r}
unique_values <- unique(df_only_delays$Hours)
unique_values <- unique(hours)
print(unique_values)
```

Formatting inorder to draw bar-plot


```{r}
df_only_delays_long <- melt(df_only_delays, id.vars = "Hours", variable.name = "variable", value.name = "value")
df_only_delays_wide <- dcast(df_only_delays_long, Hours ~ variable, value.var = "value", fun.aggregate = mean)
df_only_delays_wide <- round(df_only_delays_wide, 0)
```

Creating plot


```{r}
df_only_delays_wide$Hours <- factor(df_only_delays_wide$Hours, levels = c(0:23),
                                    labels = c("0am-1am", "1am-2am", "2am-3am", "3am-4am", "4am-5am", "5am-6am", "6am-7am", "7am-8am",
                                               "8am-9am", "9am-10am", "10am-11am", "11am-12pm", "12pm-1pm", "1pm-2pm", "2pm-3pm",
                                               "3pm-4pm", "4pm-5pm", "5pm-6pm", "6pm-7pm", "7pm-8pm", "8pm-9pm", "9pm-10pm",
                                               "10pm-11pm", "11pm-12am"))


barchart <- ggplot(df_only_delays_wide, aes(x = Hours, y = Total_Delays)) +
            geom_bar(stat = "identity", fill = "steelblue4",width = 0.8) +
            labs(title = "Average Delay Distribution by Hour of the Day",
                 x = "Time of day",
                 y = "Average Delay (min)") +
            theme(axis.text.x = element_text(angle = 43, hjust = 1))
            

            
options(repr.plot.width = 20, repr.plot.height = 7.4)



```

Printing the barchart


```{r}
print(barchart)
```

Formatting for boxplot for average delays by day of the week


```{r}


df_only_delays$DayOfWeek <- factor(df_only_delays$DayOfWeek, levels = c(1:7),
                                    labels = c("MON", "TUE", "WED", "THUR", "FRI", "SAT", "SUN"))
                                               
box_plot<-ggplot(df_only_delays, aes(x = DayOfWeek, y = Total_Delays, fill = DayOfWeek)) + ylim(c(0,100)) +
  geom_boxplot(outlier.colour="black", outlier.shape=NA, outlier.size=0) + stat_summary(fun = "mean", geom = "point", shape = 8, size = 2, color = "white")+
  labs(x = "Day of Week", y = "Total Delays")

```

Output boxplot


```{r}
print(box_plot) 
```




