---
title: "Logistic regression model fit"
author: 'Vishma Arachchi'
date: "2024-04-02"'
purpose: "FITTING A LOGISTIC REGRESSION FOR THE PROBABILITY OF DIVERTED US FLIGHTS"
output: html_document
---



IMPORTING LIBRARIES

```{r}
library(dplyr)
library(caret)
library(glmnet)
library(ROSE)
library(pROC)
library(tibble)


```


LOADING THE DATASETS


```{r}
df_a <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/2006.csv.bz2')
plane_df <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/plane-data.csv')
airport_df <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/airports.csv')

```


Replacing empty strings with NA


```{r} 

df_a[df_a == ""] <- NA
plane_df[plane_df == ""] <- NA
airport_df[airport_df == ""] <- NA

```


Creating dataframe with columns of interest


```{r}
df_main <- select(df_a, DayOfWeek, CRSDepTime,CRSArrTime, Distance, Month, UniqueCarrier, Diverted, Dest, Origin  )

```


Renaming the columns in airport dataset and merging


```{r}

colnames(airport_df)[colnames(airport_df) == "iata"] <- "Dest"
merged_df_1 <- left_join(df_main, airport_df, by = 'Dest')
colnames(merged_df_1)[colnames(merged_df_1) == "lat"] <- "Destination_latitude"
colnames(merged_df_1)[colnames(merged_df_1) == "long"] <- "Destination_longitude"
merged_df_1 <- select(merged_df_1, -c(airport, city, state, country))
colnames(airport_df)[colnames(airport_df) == "Dest"] <- "Origin"
merged_df_2 <- left_join(merged_df_1, airport_df, by = 'Origin')
colnames(merged_df_2)[colnames(merged_df_2) == "lat"] <- "Origin_latitude"
colnames(merged_df_2)[colnames(merged_df_2) == "long"] <- "Origin_longitude"
merged_df_2 <- select(merged_df_2, -c(airport, city, state, country, Dest, Origin ))

```


Converting to type factor


```{r}


month_names <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
merged_df_2$Month <- month_names[merged_df_2$Month]
merged_df_2$Month <- as.factor(merged_df_2$Month)

day_names <- c('Mon', 'Tue', 'Wed', 'Thur', 'Fri', 'Sat', 'Sun')
merged_df_2$DayOfWeek <- day_names[merged_df_2$DayOfWeek]
merged_df_2$DayOfWeek <- as.factor(merged_df_2$DayOfWeek)

merged_df_2$UniqueCarrier <- as.factor(merged_df_2$UniqueCarrier)
merged_df_2$Diverted <- as.factor(merged_df_2$Diverted)

```

Displaying class imbalance 


```{r}

summary(merged_df_2$Diverted)

```

Create a train-test split


```{r}
set.seed(123)  
train_index <- createDataPartition(merged_df_2$Diverted, p = 0.7, list = FALSE)
train_data <- merged_df_2[train_index, ]
test_data <- merged_df_2[-train_index, ]

```

Rectifying class imbalance

```{r}

undersampled_data <- downSample(train_data, train_data$Diverted  )
undersampled_data_2 <- select(undersampled_data, -Class)

```


Applying standard scaling

```{r}
train_features <- subset(undersampled_data_2, select = -Diverted)
train_target <- undersampled_data_2$Diverted

test_features <- subset(test_data, select = -Diverted)
test_target <- test_data$Diverted

```

Scaling the predictor variables (features) using the mean and standard deviation of the training data


```{r}

numerical_features_train <- subset(train_features, select = c('CRSDepTime', 'CRSArrTime', 'Distance','Destination_latitude','Destination_longitude', 'Origin_latitude','Origin_longitude'))
categorical_features_train <- subset(train_features, select = c("DayOfWeek", "Month", "UniqueCarrier"))


numerical_features_test <- subset(test_features, select = c('CRSDepTime', 'CRSArrTime', 'Distance','Destination_latitude','Destination_longitude'
                                                              ,'Origin_latitude','Origin_longitude'))

categorical_features_test <- subset(test_features, select = c("DayOfWeek", "Month", "UniqueCarrier"))

scaled_train_features <- scale(numerical_features_train)
scaled_train_features_df <- as.data.frame(scaled_train_features)
scaled_test_features <- scale(numerical_features_test, center = attr(scaled_train_features, "scaled:center"), scale = attr(scaled_train_features, "scaled:scale"))
scaled_test_features_df <- as.data.frame(scaled_test_features)


```

Applying encoding for categorical features

```{r}
categorical_encoded_data_train_model <- dummyVars("~.", data = categorical_features_train)  
categorical_encoded_data_train <- predict(categorical_encoded_data_train_model, newdata = categorical_features_train)
categorical_encoded_data_train_df <- as.data.frame(categorical_encoded_data_train)


categorical_encoded_data_test <- predict(categorical_encoded_data_train_model, newdata = categorical_features_test)
categorical_encoded_data_test_df <- as.data.frame(categorical_encoded_data_test)


```


Combining scaled features with categorical features and target variable for both train and test


```{r}
scaled_data_train <- cbind(scaled_train_features_df, categorical_encoded_data_train_df, train_target)
scaled_data_test <- cbind(scaled_test_features_df, categorical_encoded_data_test_df, test_target)


```

Renaming target to Diverted

```{r}
colnames(scaled_data_train)[colnames(scaled_data_train) == "train_target"] <- "Diverted"
colnames(scaled_data_test)[colnames(scaled_data_test) == "test_target"] <- "Diverted"


```


```{r}
summary(scaled_data_train$Diverted)


```


Fitting the model


```{r}

logistic_model <- glm(Diverted ~ ., data = scaled_data_train, family = binomial)

predicted_probabilities <- predict(logistic_model, newdata = scaled_data_test, type = 'response')
predicted_classes <- ifelse(predicted_probabilities > 0.5, 1, 0)
predicted_df <- data.frame(test_data, Predicted_Probability = predicted_probabilities, Predicted_Class = predicted_classes)


```


Drawing the roc curve

```{r}
roc_data <- roc(test_data$Diverted, predicted_probabilities) 
plot(roc_data, main = "ROC Curve for Logistic Regression", col = "blue")

legend("bottomright", legend = c("Logistic Regression"), col = c("blue"), lty = c(1, 2))


```

CONFUSION MATRIX

```{r}
conf_matrix <- confusionMatrix(data = factor(predicted_df$Predicted_Class), reference = factor(test_data$Diverted))
print(conf_matrix)
```

Visualizing the Coeffitients
```{r}
coefficients <- coef(logistic_model)
coefficient_df <- data.frame(coefficients)
coefficient_df$coefficients <- replace(coefficient_df$coefficients, is.na(coefficient_df$coefficients), 0)
coefficient_df <- rownames_to_column(coefficient_df, var = "IndexColumn")

ggplot(coefficient_df, aes(x = IndexColumn, y = coefficients)) +
  geom_bar(stat = "identity", fill = ifelse(coefficient_df$coefficients > 0, "blue", "red")) +
  labs(title = "Coefficients Visualization", x = "Features", y = "Coefficient Value") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  

```


