---
title: "DO OLDER PLANES SUFFER MORE DELAYS?"
author: 'Vishma Arachchi'
date: "2024-04-02"
purpose: "To explore the relationship between plane age and corresponding delay times"
output: html_document
---



LOADING LIBRARIES

```{r}
library(ggplot2)
library(repr)
library(tidyr)
library(dplyr)
library(reshape2)

```

Loading the datasets

```{r}
df_com <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/2006.csv.bz2')
plane_df <- read.csv('/Users/vishmaarachchi/Documents/BSC/Programming coursework - PROJECT/plane-data.csv')

```

Replacing null values

```{r}
df_com[df_com == ""] <- NA
plane_df[plane_df == ""] <- NA

```

Conversion to type numeric

```{r}
df_com$DepDelay <- as.numeric(df_com$DepDelay)
df_com$ArrDelay <- as.numeric(df_com$ArrDelay)
df_com$CRSDepTime <- as.character(df_com$CRSDepTime)

```

Calculating the mean of ArrDelayand DepDelay

```{r}
mean_ArrDelay <- mean(df_com$ArrDelay,na.rm = TRUE)
mean_DepDelay <- mean(df_com$DepDelay,na.rm = TRUE)

```

Replacing NA values

```{r}

df_com$ArrDelay[is.na(df_com$ArrDelay)] <- round(mean_ArrDelay,1)
df_com$DepDelay[is.na(df_com$DepDelay)] <- round(mean_DepDelay,1)

```

Removing rows of plane dataset with Year of Service commencement is 0 or NONE

```{r}

colnames(plane_df)[colnames(plane_df) == "year"] <- "Start_of_Service"
colnames(plane_df)[colnames(plane_df) == "tailnum"] <- "TailNum"


plane_df <- plane_df[complete.cases(plane_df$Start_of_Service), ]
plane_df <- subset(plane_df, Start_of_Service != '0000')
plane_df <- subset(plane_df, Start_of_Service != "None")

plane_df$Start_of_Service = as.numeric(plane_df$Start_of_Service)

```

Calculating Correlation coeffitient between ArrDelay and DepDelay

```{r}

correlation_coefficient <- cor(df_com$ArrDelay, df_com$DepDelay)
print(correlation_coefficient)

```

Creating a new column for total delays in combined dataframe

```{r}
df_com$Total_Delays <- df_com$ArrDelay + df_com$DepDelay

```

New dataframe with relevant columns

```{r}
df_com_1 <- select(df_com, Year, TailNum,Total_Delays )
```

Merging the two datasets and filtering

```{r}
merged_df <- merge.data.frame(df_com_1, plane_df, by = 'TailNum',all.x = TRUE)
merged_df_2 <- select(merged_df, Year, TailNum,Start_of_Service,Total_Delays )

```

Dataframe with only datapoints where overall delay occurs

```{r}
df_only_delays <- subset(merged_df_2, Total_Delays > 0)

```

Removing Start of Service year = NA rows

```{r}

df_only_delays <- df_only_delays[complete.cases(df_only_delays$Start_of_Service), ]

```

Calculating the ages of each plane and removing all negative ages

```{r}
df_only_delays$Age <- df_only_delays$Year - df_only_delays$Start_of_Service
df_only_delays <- subset(df_only_delays, Age != -1)


```

Drawing a scatter plot

1)Formatting data for a scatter plot

```{r}

select_variable <- c("Total_Delays")

df_only_delays_long <- melt(df_only_delays, id.vars = "Age", measure.vars = select_variable, value.name = "value")

df_only_delays_wide <- df_only_delays_long %>%
  group_by(Age, variable) %>%
  summarise(value = mean(value, na.rm = TRUE)) %>%
  spread(key = variable, value = value)


```

```{r}
ggplot(df_only_delays_wide, aes(x = Age, y = Total_Delays)) +
  geom_point() +  # 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Scatterplot with Regression Line", x = "Age of Plane (Years)", y = "Average Delay (min)")


```

Correlation between ages and average delay

```{r}
print(cor(df_only_delays_wide$Age,df_only_delays_wide$Total_Delays))


```

Visualizing a boxplot with age groups as classes

a)  Data preparation

```{r}


values_to_be_mapped_list <- list(c(0:11), c(12:21), c(22:31), c(32:41), c(42:51))
replacement_value_list <- c('0-11', '12-21', '22-31', '32-41', '42-51')


df_only_delays$Age_groups <- sapply(df_only_delays$Age, function(x) {
  for (i in 1:length(values_to_be_mapped_list)) {
    if (x %in% values_to_be_mapped_list[[i]]) {
      return(replacement_value_list[i])
    }
  }
})


```

b)Drawing the boxplot


```{r}
df_only_delays$Age_groups <- as.character(df_only_delays$Age_groups) # Convert factor to character
df_only_delays$Age_groups[is.na(df_only_delays$Age_groups)] <- "Unknown"



box_plot<-ggplot(df_only_delays, aes(x = Age_groups, y = Total_Delays, fill = Age_groups)) + ylim(c(0,300)) +
  geom_boxplot(outlier.colour="black", outlier.shape=NA, outlier.size=0) + stat_summary(fun = "mean", geom = "point", shape = 8,
                                                                                        size = 2, color = "white")+
  labs(x = "Age Groups", y = "Average Delay")

box_plot
```


